The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities

### 2024-03-05 r_r/submission/manuscript_rihe_r_r.Rmd
copy edits abstract through methods

Scholarship on nonresident enrollment by public research universities has developed in isolation from scholarship on linkages between private high schools and selective private universities. We argue that these literatures are part of a broader story about the competition for students from affluent schools and communities. This manuscript analyzes off-campus recruiting visits to private high schools made by a convenience sample of 15 public research universities and 14 selective private universities. An off-campus recruiting visit indicates a social relationship between a school and a university. Therefore, we utilize social network methods to examine the recruiting networks of public and private universities. With respect to scale (research question 1), universities in our sample made a disproportionate number of visits to private high schools. With respect to overlap (RQ2), simple network analyses and community detection methods reveal substantial overlap in the recruiting networks of public and private universities. RQ3 assesses the characteristics of visited schools. Both public and private universities tended to visit private schools in their home geographic region and also in the South, where private school enrollment growth has been strongest. Visited private schools enroll a much larger share of white students than visited public schools. Surprisingly, several public research universities visited sectarian private high schools at a rate similar to sectarian private universities.

@@

Autumn is college recruiting season at Harpeth Hall, a nonsectarian private school for girls located in the affluent Green Hills neighborhood of Nashville, Tennessee. Founded in 1951, Harpeth is an "A+" school ranked 337 in the nation according to the 2020 *Best Private High Schools in America* by Niche. Enrollment patterns at Nashville-area private schools reflect broader regional patterns, whereby "efforts by the courts to desegregate public schools in the South resulted in White flight to private schools" [@RN4504, p. 14]. In Fall 2017, about 85% of Harpeth students were white and less than 5% were Black or Latinx (Author calculations).

During the 2017 calendar year, we tracked off-campus recruiting visits made by 14 selective private universities, 12 selective private liberal arts colleges, and 15 public research universities.^[These institutions are: Northwestern, Notre Dame, Emory, Tufts, Boston College, Tulane, Case Western Reserve, Villanova, Southern Methodist University, Baylor, University of Denver, Texas Christian University, Stevens Institute of Technology, Marquette, Williams, Swarthmore, Middlebury, Smith, Harvey Mudd, Colorado College, Macalester, Scripps, Oberlin, Occidental, Sewanee, Connecticut College, UC-Berkeley, UC-San Diego, University of Georgia, University of Pittsburgh, Rutgers, University of Massachusetts-Amherst, UC-Riverside, SUNY Stony Brook, University of Colorado-Boulder, University of South Carolina, University of Kansas, University of Nebraska-Lincoln, University of Alabama, University of Cincinnati, and University of Arkansas.] From September 6th through October 17th, Harpeth Hall received recruiting visits from 11 private universities in our sample (Northwestern, Boston College, Tufts, Villanova, Case Western Reserve, University of Denver, Emory, Tulane, Baylor, Texas Christian University, Southern Methodist University), and three private liberal arts colleges (Harvey Mudd, Scripps, and Sewanee). Previous scholarship observes strong connections between private schools and selective private universities [@RN3519; @RN4407]. However, Harpeth also received visits from six out-of-state public research universities, the University of Cincinnati, University of South Carolina, the University of Alabama, University of Arkansas, University of Georgia, and CU Boulder.

The nonsectarian University School of Nashville was founded in 1975 and is located next door to Vanderbilt University. It is an "A+" school, ranked 127th in the nation, and is more racially diverse than Harpeth (64% white, 11.5% Black, 3.5% Latinx, 10.0% Asian). From August 22nd 2017 through October 18th, the University School received visits from nearly the same set of 11 private universities that visited Harpeth (Notre Dame visited and Baylor did not) and five liberal arts colleges (including Williams and Middlebury), but only two public research universities from our sample (University of Cincinnati, CU Boulder).

@@

The results yield practical implications for public and private universities concerned about racial and socioeconomic diversity. Following the Covid-19 Pandemic, off-campus recruiting visits have bounced back because they remain a fundamental means of maintaining relationships with high school guidance counselors [@RN4896; @RN4895; @eab_nacac_counselor_2023]. Prior research finds that students are sensitive to recruiting interventions, including off-campus recruiting visits [@RN4324; @RN4402]. Therefore, which high schools a university chooses to visit affects the subsequent enrollment composition of the university. In the aftermath of the Supreme Court decision on race-conscious admissions, public and private universities are searching for ways to increase racial diversity without running afoul of the law. The @RN4954[p. 9] recommends that universities expand "outreach to high schools with substantial populations of low-income students and students of color." Most public and private universities in our sample visited a disproportionate number of private schools and these private schools tended to enroll mostly white students. We recommend that universities concerned about racial diversity should conduct an internal equity audit of their recruiting practices, for example, comparing the racial and socioeconomic characteristics of schools they visit versus schools they do not visit.

@@

<!-- PREVIOUS PARAGRAPH
We situate this study vis-a-vis scholarship on nonresident enrollment at public research universities, which lies within the broader literature on "privatization" [CITE]. Against the backdrop of many states disinvesting in higher education, the privatization literature examines the ways that public universities start behaving like private universities by emphasizing revenue generation and the pursuit of prestige in research and enrollment management behaviors. The so-called "iron triangle" of enrollment management states that universities pursue the three broad enrollment goals of access, academic profile, and revenue generation [CITE]. "Because resources are scarce, the imagery of the iron triangle suggests that the pursuit of one enrollment goal may involve tradeoffs with other goals. Scholarship on privatization expressed concern that the public research universities were emphasizing tuition revenue and rankings and deemphasizing access for state residents.-->

@@

Other studies of nonresident enrollment examine university enrollment composition. Using data from NPSAS 2011-12, @RN3685 show that compared to resident students at public research universities, nonresident students generate higher net tuition revenue, are more affluent, are less likely to be Pell recipients, and are less likely to be Black or LatinX. Institution-level panel models reveal that growth in the share of nonresident students was associated with a decline in the share of Pell recipients and a decline in the share of Black and LatinX students. @RN4290 examine whether growth in the number of nonresident students causes resident enrollment to decline. They find that nonresident enrollment "crowds out" resident enrollment at prestigious public research universities, but not at less prestigious public research universities.

@@

@RN4504<!-- use data from several national surveys to --> set the stage by describing long-term trends in private school enrollment. The percentage of American (elementary school) students attending private schools declined from 15% in the mid-1960s to 10% in the mid-1970s, and declined gradually from 11% in 1999 to 9% in 2015. The private school enrollment rate for high-income families and for low-income families has remained relatively stable over time at about 17% and 4%, respectively. The enrollment rate for middle-income families declined from 12% in 1968 to 6% in 2013. This decline was driven by shrinking Catholic school enrollment, which represented 89% of private school enrollment in 1965 and 42% of enrollment in 2013. Moreover, contemporary Catholic school students tend to be more affluent compared to prior decades. There has been a long-term increase in private school enrollment in the South, with middle-income families increasingly attending Christian schools and high-income families increasingly attending nonsectarian schools. With respect to race, 11% of white students attended private school in 2013, compared to 5% of Black students, and 3% of Hispanic students.
<!-- Private school enrollment rates are higher for families located in urban areas than those in suburban areas.
 However, for high income families, rates of private school enrollment do not differ substantially across race. -->

@@

The High School Longitudinal Survey of 2009 (HSLS) tracks the postsecondary destinations of students who were in ninth grade in 2009. Simple descriptive statistics are eye-opening (Author calculations).^[Analyses consist of students who met the following conditions: completed the 2012 follow-up survey; obtained high school transcript data; and completed the 2016 follow-up survey. The survey weight variable used was W4W1W2W3STU. After weighting, these 14,020 (unweighted) students represent a population of 3.93 million U.S. ninth graders in 2009. Data on college rankings come from the 2020 U.S. News \& World Report *Best Colleges*. Data on high school control come from the HSLS school file and from the Private School Survey.] About 93% of students attended a public high school and 7% attended a private high school, including 3.6% attending a Catholic school, 1.6% attending a nonsectarian school, and 1.8% attending a Christian school. Of public high school students, 5.0% attending a public university ranked in the top 100 by U.S. News and World Report (USNWR) and 1.9% attended a private university ranked in the top 100. Of private high school students, 10.9% attended a top 100 public university and 16.8% attended a top 100 private university. For Catholic school students, 12.8% attended a top 100 public university and 11.7% attended a top 100 private university, while for Christian schools these percentages were 8.2% and 8.2%<!--TODO: Want to double check they're both 8.2%?-->, respectively. However, for nonsectarian private school students, 11.3% attended a top 100 public university and an incredible 39.8% attended a top 100 private university.

Focusing on the institutional composition of private universities in USNWR top 100, 39.8%<!--TODO: Want to double check this and the previous sentence are both 39.8%?--> attended a private high school (compared to 7% of all high school students who attended a private high school) and 21.2% attended a nonsectarian private high school (compared to 1.6% of all high school students who attended a nonsectarian private high school).

@@

Analyses utilized several secondary data sources. The data on university characteristics come from the 2017 Integrated Postsecondary Education Data System (IPEDS) and 2020 Best Colleges ranking by U.S. News & World Report. For the high schools in our analysis, we use secondary data from the 2017-18 NCES Common Core of Data (CCD) for public high schools, and the 2017-18 NCES Private School Universe Survey (PSS) and 2020 Best Private High Schools ranking by Niche for private high schools.^[For both public and private high schools, we also used NCES data from previous years when the 2017-18 data was not available for a small number of schools.]

@@

This article faces several limitations. First, our data collection sample consisted of public research-extensive universities as defined by the 2000 Carnegie Classification (N=102) and all private universities in the top 100 of U.S. News and World Report National Universities rankings (N=57). Making comparisons between public and private universities would be more appropriate if we used the same classification method to define the data collection sample for all universities. The process we used to define the data collection sample yielded a sample of public research universities with lower USNWR rankings than the analysis sample of selective private universities. However, the columns in Table \@ref(tab:sample-characteristics) for 25th and 75th percentile SAT score of enrolled freshmen suggest that private universities ranked lower than 75 have test score ranges that are fairly similar to public research universities not ranked in the top 100.

>>>>

Scholarship on nonresident enrollment by public research universities has developed in isolation from scholarship on linkages between private high schools and selective private universities. We argue that these literatures are part of a broader story about the competition for students from affluent schools and communities. This manuscript analyzes off-campus recruiting visits to private high schools made by a convenience sample of 15 public research universities and 14 selective private universities. An off-campus recruiting visit indicates a social relationship between a school and a university. Therefore, we utilize social network methods to examine the recruiting networks of public and private universities. With respect to scale (research question (RQ) 1), universities in our sample made a disproportionate number of visits to private high schools. With respect to overlap (RQ2), simple network analyses and community detection methods reveal substantial overlap in the recruiting networks of public and private universities. RQ3 assesses the characteristics of visited schools. Both public and private universities tended to visit private schools in their home geographic region and also in the South, where private school enrollment growth has been strongest. Visited private schools enroll a much larger share of white students than visited public schools. Surprisingly, several public research universities visited sectarian private high schools at a rate similar to sectarian private universities.

@@

Autumn is college recruiting season at Harpeth Hall, a nonsectarian private school for girls located in the affluent Green Hills neighborhood of Nashville, Tennessee. Founded in 1951, Harpeth is an "A+" school ranked 337 in the nation according to the 2020 *Best Private High Schools in America* by Niche. Enrollment patterns at Nashville-area private schools reflect broader regional patterns, whereby "efforts by the courts to desegregate public schools in the South resulted in White flight to private schools" [@RN4504, p. 14]. In Fall 2017, about 85% of Harpeth Hall students were white and less than 5% were Black or Latinx (Author calculations).

During the 2017 calendar year, we tracked off-campus recruiting visits made by 14 selective private universities, 12 selective private liberal arts colleges, and 15 public research universities.^[These institutions are: Northwestern, Notre Dame, Emory, Tufts, Boston College, Tulane, Case Western Reserve, Villanova, Southern Methodist University, Baylor, University of Denver, Texas Christian University, Stevens Institute of Technology, Marquette, Williams, Swarthmore, Middlebury, Smith, Harvey Mudd, Colorado College, Macalester, Scripps, Oberlin, Occidental, Sewanee, Connecticut College, UC-Berkeley, UC-San Diego, University of Georgia, University of Pittsburgh, Rutgers, University of Massachusetts-Amherst, UC-Riverside, SUNY Stony Brook, University of Colorado-Boulder, University of South Carolina, University of Kansas, University of Nebraska-Lincoln, University of Alabama, University of Cincinnati, and University of Arkansas.] From September 6th through October 17th, Harpeth Hall received recruiting visits from 11 private universities in our sample (Northwestern, Boston College, Tufts, Villanova, Case Western Reserve, University of Denver, Emory, Tulane, Baylor, Texas Christian University, Southern Methodist University), and three private liberal arts colleges (Harvey Mudd, Scripps, and Sewanee). Previous scholarship observes strong connections between private schools and selective private universities [@RN3519; @RN4407]. However, Harpeth Hall also received visits from six out-of-state public research universities, the University of Cincinnati, University of South Carolina, the University of Alabama, University of Arkansas, University of Georgia, and CU Boulder.

The nonsectarian University School of Nashville was founded in 1975 and is located next door to Vanderbilt University. It is an "A+" school, ranked 127th in the nation, and is more racially diverse than Harpeth Hall (64% white, 11.5% Black, 3.5% Latinx, 10.0% Asian). From August 22nd 2017 through October 18th, the University School of Nashville received visits from nearly the same set of 11 private universities that visited Harpeth Hall (Notre Dame visited and Baylor did not) and five liberal arts colleges (including Williams and Middlebury), but only two public research universities from our sample (University of Cincinnati, CU Boulder).

@@

The results yield practical implications for public and private universities concerned about racial and socioeconomic diversity. Following the COVID-19 Pandemic, off-campus recruiting visits have bounced back because they remain a fundamental means of maintaining relationships with high school guidance counselors [@RN4896; @RN4895; @eab_nacac_counselor_2023]. Prior research finds that students are sensitive to recruiting interventions, including off-campus recruiting visits [@RN4324; @RN4402]. Therefore, which high schools a university chooses to visit affects the subsequent enrollment composition of the university. In the aftermath of the Supreme Court decision on race-conscious admissions, public and private universities are searching for ways to increase racial diversity without running afoul of the law. The @RN4954[p. 9] recommends that universities expand "outreach to high schools with substantial populations of low-income students and students of color." Most public and private universities in our sample visited a disproportionate number of private schools and these private schools tended to enroll mostly white students. We recommend that universities concerned about racial diversity should conduct an internal equity audit of their recruiting practices, for example, comparing the racial and socioeconomic characteristics of schools they visit versus schools they do not visit.

@@

<!-- PREVIOUS PARAGRAPH
We situate this study vis-a-vis scholarship on nonresident enrollment at public research universities, which lies within the broader literature on "privatization" [CITE]. Against the backdrop of many states disinvesting in higher education, the privatization literature examines the ways that public universities start behaving like private universities by emphasizing revenue generation and the pursuit of prestige in research and enrollment management behaviors. The so-called "iron triangle" of enrollment management states that universities pursue the three broad enrollment goals of access, academic profile, and revenue generation [CITE]. "Because resources are scarce, the imagery of the iron triangle suggests that the pursuit of one enrollment goal may involve trade-offs with other goals. Scholarship on privatization expressed concern that the public research universities were emphasizing tuition revenue and rankings and de-emphasizing access for state residents.-->

@@

Other studies of nonresident enrollment examine university enrollment composition. Using data from NPSAS 2011-12, @RN3685 show that compared to resident students at public research universities, nonresident students generate higher net tuition revenue, are more affluent, are less likely to be Pell recipients, and are less likely to be Black or Latinx. Institution-level panel models reveal that growth in the share of nonresident students was associated with a decline in the share of Pell recipients and a decline in the share of Black and Latinx students. @RN4290 examine whether growth in the number of nonresident students causes resident enrollment to decline. They find that nonresident enrollment "crowds out" resident enrollment at prestigious public research universities, but not at less prestigious public research universities.

@@

@RN4504<!-- use data from several national surveys to --> set the stage by describing long-term trends in private school enrollment. The percentage of American (elementary school) students attending private schools declined from 15% in the mid-1960s to 10% in the mid-1970s, and declined gradually from 11% in 1999 to 9% in 2015. The private school enrollment rate for high-income families and for low-income families has remained relatively stable over time at about 17% and 4%, respectively. The enrollment rate for middle-income families declined from 12% in 1968 to 6% in 2013. This decline was driven by shrinking Catholic school enrollment, which represented 89% of private school enrollment in 1965 and 42% of enrollment in 2013. Moreover, contemporary Catholic school students tend to be more affluent compared to prior decades. There has been a long-term increase in private school enrollment in the South, with middle-income families increasingly attending Christian schools and high-income families increasingly attending nonsectarian schools. With respect to race, 11% of white students attended private school in 2013, compared to 5% of Black students and 3% of Hispanic students.
<!-- Private school enrollment rates are higher for families located in urban areas than those in suburban areas.
 However, for high income families, rates of private school enrollment do not differ substantially across race. -->

@@

The High School Longitudinal Survey of 2009 (HSLS) tracks the postsecondary destinations of students who were in ninth grade in 2009. Simple descriptive statistics are eye-opening (Author calculations).^[Analyses consist of students who met the following conditions: completed the 2012 follow-up survey; obtained high school transcript data; and completed the 2016 follow-up survey. The survey weight variable used was W4W1W2W3STU. After weighting, these 14,020 (unweighted) students represent a population of 3.93 million U.S. ninth graders in 2009. Data on college rankings come from the 2020 U.S. News \& World Report *Best Colleges*. Data on high school control come from the HSLS school file and from the Private School Survey.] About 93% of students attended a public high school and 7% attended a private high school, including 3.6% attending a Catholic school, 1.6% attending a nonsectarian school, and 1.8% attending a Christian school. Of public high school students, 5.0% attending a public university ranked in the top 100 by U.S. News and World Report and 1.9% attended a private university ranked in the top 100. Of private high school students, 10.9% attended a top 100 public university and 16.8% attended a top 100 private university. For Catholic school students, 12.8% attended a top 100 public university and 11.7% attended a top 100 private university, while for Christian schools these percentages were 8.2% and 8.2%<!--TODO: Want to double check they're both 8.2%?-->, respectively. However, for nonsectarian private school students, 11.3% attended a top 100 public university and an incredible 39.8% attended a top 100 private university.

Focusing on the institutional composition of private universities in U.S. News and World Report's top 100, 39.8%<!--TODO: Want to double check this and the previous sentence are both 39.8%?--> attended a private high school (compared to 7% of all high school students who attended a private high school) and 21.2% attended a nonsectarian private high school (compared to 1.6% of all high school students who attended a nonsectarian private high school).

@@

Analyses utilized several secondary data sources. The data on university characteristics come from the 2017 Integrated Postsecondary Education Data System (IPEDS) and 2020 Best Colleges ranking by U.S. News and World Report. For the high schools in our analysis, we use secondary data from the 2017-18 NCES Common Core of Data (CCD) for public high schools, and the 2017-18 NCES Private School Universe Survey (PSS) and 2020 Best Private High Schools ranking by Niche for private high schools.^[For both public and private high schools, we also used NCES data from previous years when the 2017-18 data was not available for a small number of schools.]

@@

This article faces several limitations. First, our data collection sample consisted of public research-extensive universities as defined by the 2000 Carnegie Classification (N=102) and all private universities in the top 100 of U.S. News and World Report National Universities rankings (N=57). Making comparisons between public and private universities would be more appropriate if we used the same classification method to define the data collection sample for all universities. The process we used to define the data collection sample yielded a sample of public research universities with lower U.S. News and World Report rankings than the analysis sample of selective private universities. However, the columns in Table \@ref(tab:sample-characteristics) for 25th and 75th percentile SAT score of enrolled freshmen suggest that private universities ranked lower than 75 have test score ranges that are fairly similar to public research universities not ranked in the top 100.

### 2024-03-05 r_r/submission/manuscript_rihe_r_r.tex
copy edits abstract through methods

Autumn is college recruiting season at Harpeth Hall, a nonsectarian private school for girls located in the affluent Green Hills neighborhood of Nashville, Tennessee. Founded in 1951, Harpeth is an ``A+'' school ranked 337 in the nation according to the 2020 \emph{Best Private High Schools in America} by Niche. Enrollment patterns at Nashville-area private schools reflect broader regional patterns, whereby ``efforts by the courts to desegregate public schools in the South resulted in White flight to private schools'' (Murnane \& Reardon, 2018, p. 14). In Fall 2017, about 85\% of Harpeth students were white and less than 5\% were Black or Latinx (Author calculations).

During the 2017 calendar year, we tracked off-campus recruiting visits made by 14 selective private universities, 12 selective private liberal arts colleges, and 15 public research universities.\footnote{These institutions are: Northwestern, Notre Dame, Emory, Tufts, Boston College, Tulane, Case Western Reserve, Villanova, Southern Methodist University, Baylor, University of Denver, Texas Christian University, Stevens Institute of Technology, Marquette, Williams, Swarthmore, Middlebury, Smith, Harvey Mudd, Colorado College, Macalester, Scripps, Oberlin, Occidental, Sewanee, Connecticut College, UC-Berkeley, UC-San Diego, University of Georgia, University of Pittsburgh, Rutgers, University of Massachusetts-Amherst, UC-Riverside, SUNY Stony Brook, University of Colorado-Boulder, University of South Carolina, University of Kansas, University of Nebraska-Lincoln, University of Alabama, University of Cincinnati, and University of Arkansas.} From September 6th through October 17th, Harpeth Hall received recruiting visits from 11 private universities in our sample (Northwestern, Boston College, Tufts, Villanova, Case Western Reserve, University of Denver, Emory, Tulane, Baylor, Texas Christian University, Southern Methodist University), and three private liberal arts colleges (Harvey Mudd, Scripps, and Sewanee). Previous scholarship observes strong connections between private schools and selective private universities (Khan, 2011; Stevens, 2007). However, Harpeth also received visits from six out-of-state public research universities, the University of Cincinnati, University of South Carolina, the University of Alabama, University of Arkansas, University of Georgia, and CU Boulder.

The nonsectarian University School of Nashville was founded in 1975 and is located next door to Vanderbilt University. It is an ``A+'' school, ranked 127th in the nation, and is more racially diverse than Harpeth (64\% white, 11.5\% Black, 3.5\% Latinx, 10.0\% Asian). From August 22nd 2017 through October 18th, the University School received visits from nearly the same set of 11 private universities that visited Harpeth (Notre Dame visited and Baylor did not) and five liberal arts colleges (including Williams and Middlebury), but only two public research universities from our sample (University of Cincinnati, CU Boulder).

@@

The results yield practical implications for public and private universities concerned about racial and socioeconomic diversity. Following the Covid-19 Pandemic, off-campus recruiting visits have bounced back because they remain a fundamental means of maintaining relationships with high school guidance counselors (EAB \& NACAC, 2023; Ruffalo Noel-Levitz, 2022a, 2022b). Prior research finds that students are sensitive to recruiting interventions, including off-campus recruiting visits (Holland, 2019; Ruffalo Noel-Levitz, 2018). Therefore, which high schools a university chooses to visit affects the subsequent enrollment composition of the university. In the aftermath of the Supreme Court decision on race-conscious admissions, public and private universities are searching for ways to increase racial diversity without running afoul of the law. The U.S. Department of Education (2023, p. 9) recommends that universities expand ``outreach to high schools with substantial populations of low-income students and students of color.'' Most public and private universities in our sample visited a disproportionate number of private schools and these private schools tended to enroll mostly white students. We recommend that universities concerned about racial diversity should conduct an internal equity audit of their recruiting practices, for example, comparing the racial and socioeconomic characteristics of schools they visit versus schools they do not visit.

@@

Other studies of nonresident enrollment examine university enrollment composition. Using data from NPSAS 2011-12, Jaquette et al. (2016) show that compared to resident students at public research universities, nonresident students generate higher net tuition revenue, are more affluent, are less likely to be Pell recipients, and are less likely to be Black or LatinX. Institution-level panel models reveal that growth in the share of nonresident students was associated with a decline in the share of Pell recipients and a decline in the share of Black and LatinX students. Curs \& Jaquette (2017) examine whether growth in the number of nonresident students causes resident enrollment to decline. They find that nonresident enrollment ``crowds out'' resident enrollment at prestigious public research universities, but not at less prestigious public research universities.

@@

Murnane \& Reardon (2018) set the stage by describing long-term trends in private school enrollment. The percentage of American (elementary school) students attending private schools declined from 15\% in the mid-1960s to 10\% in the mid-1970s, and declined gradually from 11\% in 1999 to 9\% in 2015. The private school enrollment rate for high-income families and for low-income families has remained relatively stable over time at about 17\% and 4\%, respectively. The enrollment rate for middle-income families declined from 12\% in 1968 to 6\% in 2013. This decline was driven by shrinking Catholic school enrollment, which represented 89\% of private school enrollment in 1965 and 42\% of enrollment in 2013. Moreover, contemporary Catholic school students tend to be more affluent compared to prior decades. There has been a long-term increase in private school enrollment in the South, with middle-income families increasingly attending Christian schools and high-income families increasingly attending nonsectarian schools. With respect to race, 11\% of white students attended private school in 2013, compared to 5\% of Black students, and 3\% of Hispanic students.

@@

The High School Longitudinal Survey of 2009 (HSLS) tracks the postsecondary destinations of students who were in ninth grade in 2009. Simple descriptive statistics are eye-opening (Author calculations).\footnote{Analyses consist of students who met the following conditions: completed the 2012 follow-up survey; obtained high school transcript data; and completed the 2016 follow-up survey. The survey weight variable used was W4W1W2W3STU. After weighting, these 14,020 (unweighted) students represent a population of 3.93 million U.S. ninth graders in 2009. Data on college rankings come from the 2020 U.S. News \& World Report \emph{Best Colleges}. Data on high school control come from the HSLS school file and from the Private School Survey.} About 93\% of students attended a public high school and 7\% attended a private high school, including 3.6\% attending a Catholic school, 1.6\% attending a nonsectarian school, and 1.8\% attending a Christian school. Of public high school students, 5.0\% attending a public university ranked in the top 100 by U.S. News and World Report (USNWR) and 1.9\% attended a private university ranked in the top 100. Of private high school students, 10.9\% attended a top 100 public university and 16.8\% attended a top 100 private university. For Catholic school students, 12.8\% attended a top 100 public university and 11.7\% attended a top 100 private university, while for Christian schools these percentages were 8.2\% and 8.2\%, respectively. However, for nonsectarian private school students, 11.3\% attended a top 100 public university and an incredible 39.8\% attended a top 100 private university.

Focusing on the institutional composition of private universities in USNWR top 100, 39.8\% attended a private high school (compared to 7\% of all high school students who attended a private high school) and 21.2\% attended a nonsectarian private high school (compared to 1.6\% of all high school students who attended a nonsectarian private high school).

@@

Analyses utilized several secondary data sources. The data on university characteristics come from the 2017 Integrated Postsecondary Education Data System (IPEDS) and 2020 Best Colleges ranking by U.S. News \& World Report. For the high schools in our analysis, we use secondary data from the 2017-18 NCES Common Core of Data (CCD) for public high schools, and the 2017-18 NCES Private School Universe Survey (PSS) and 2020 Best Private High Schools ranking by Niche for private high schools.\footnote{For both public and private high schools, we also used NCES data from previous years when the 2017-18 data was not available for a small number of schools.}

@@

This article faces several limitations. First, our data collection sample consisted of public research-extensive universities as defined by the 2000 Carnegie Classification (N=102) and all private universities in the top 100 of U.S. News and World Report National Universities rankings (N=57). Making comparisons between public and private universities would be more appropriate if we used the same classification method to define the data collection sample for all universities. The process we used to define the data collection sample yielded a sample of public research universities with lower USNWR rankings than the analysis sample of selective private universities. However, the columns in Table \ref{tab:sample-characteristics} for 25th and 75th percentile SAT score of enrolled freshmen suggest that private universities ranked lower than 75 have test score ranges that are fairly similar to public research universities not ranked in the top 100.

>>>>

Autumn is college recruiting season at Harpeth Hall, a nonsectarian private school for girls located in the affluent Green Hills neighborhood of Nashville, Tennessee. Founded in 1951, Harpeth is an ``A+'' school ranked 337 in the nation according to the 2020 \emph{Best Private High Schools in America} by Niche. Enrollment patterns at Nashville-area private schools reflect broader regional patterns, whereby ``efforts by the courts to desegregate public schools in the South resulted in White flight to private schools'' (Murnane \& Reardon, 2018, p. 14). In Fall 2017, about 85\% of Harpeth Hall students were white and less than 5\% were Black or Latinx (Author calculations).

During the 2017 calendar year, we tracked off-campus recruiting visits made by 14 selective private universities, 12 selective private liberal arts colleges, and 15 public research universities.\footnote{These institutions are: Northwestern, Notre Dame, Emory, Tufts, Boston College, Tulane, Case Western Reserve, Villanova, Southern Methodist University, Baylor, University of Denver, Texas Christian University, Stevens Institute of Technology, Marquette, Williams, Swarthmore, Middlebury, Smith, Harvey Mudd, Colorado College, Macalester, Scripps, Oberlin, Occidental, Sewanee, Connecticut College, UC-Berkeley, UC-San Diego, University of Georgia, University of Pittsburgh, Rutgers, University of Massachusetts-Amherst, UC-Riverside, SUNY Stony Brook, University of Colorado-Boulder, University of South Carolina, University of Kansas, University of Nebraska-Lincoln, University of Alabama, University of Cincinnati, and University of Arkansas.} From September 6th through October 17th, Harpeth Hall received recruiting visits from 11 private universities in our sample (Northwestern, Boston College, Tufts, Villanova, Case Western Reserve, University of Denver, Emory, Tulane, Baylor, Texas Christian University, Southern Methodist University), and three private liberal arts colleges (Harvey Mudd, Scripps, and Sewanee). Previous scholarship observes strong connections between private schools and selective private universities (Khan, 2011; Stevens, 2007). However, Harpeth Hall also received visits from six out-of-state public research universities, the University of Cincinnati, University of South Carolina, the University of Alabama, University of Arkansas, University of Georgia, and CU Boulder.

The nonsectarian University School of Nashville was founded in 1975 and is located next door to Vanderbilt University. It is an ``A+'' school, ranked 127th in the nation, and is more racially diverse than Harpeth Hall (64\% white, 11.5\% Black, 3.5\% Latinx, 10.0\% Asian). From August 22nd 2017 through October 18th, the University School of Nashville received visits from nearly the same set of 11 private universities that visited Harpeth Hall (Notre Dame visited and Baylor did not) and five liberal arts colleges (including Williams and Middlebury), but only two public research universities from our sample (University of Cincinnati, CU Boulder).

@@

The results yield practical implications for public and private universities concerned about racial and socioeconomic diversity. Following the COVID-19 Pandemic, off-campus recruiting visits have bounced back because they remain a fundamental means of maintaining relationships with high school guidance counselors (EAB \& NACAC, 2023; Ruffalo Noel-Levitz, 2022a, 2022b). Prior research finds that students are sensitive to recruiting interventions, including off-campus recruiting visits (Holland, 2019; Ruffalo Noel-Levitz, 2018). Therefore, which high schools a university chooses to visit affects the subsequent enrollment composition of the university. In the aftermath of the Supreme Court decision on race-conscious admissions, public and private universities are searching for ways to increase racial diversity without running afoul of the law. The U.S. Department of Education (2023, p. 9) recommends that universities expand ``outreach to high schools with substantial populations of low-income students and students of color.'' Most public and private universities in our sample visited a disproportionate number of private schools and these private schools tended to enroll mostly white students. We recommend that universities concerned about racial diversity should conduct an internal equity audit of their recruiting practices, for example, comparing the racial and socioeconomic characteristics of schools they visit versus schools they do not visit.

@@

Other studies of nonresident enrollment examine university enrollment composition. Using data from NPSAS 2011-12, Jaquette et al. (2016) show that compared to resident students at public research universities, nonresident students generate higher net tuition revenue, are more affluent, are less likely to be Pell recipients, and are less likely to be Black or Latinx. Institution-level panel models reveal that growth in the share of nonresident students was associated with a decline in the share of Pell recipients and a decline in the share of Black and Latinx students. Curs \& Jaquette (2017) examine whether growth in the number of nonresident students causes resident enrollment to decline. They find that nonresident enrollment ``crowds out'' resident enrollment at prestigious public research universities, but not at less prestigious public research universities.

@@

Murnane \& Reardon (2018) set the stage by describing long-term trends in private school enrollment. The percentage of American (elementary school) students attending private schools declined from 15\% in the mid-1960s to 10\% in the mid-1970s, and declined gradually from 11\% in 1999 to 9\% in 2015. The private school enrollment rate for high-income families and for low-income families has remained relatively stable over time at about 17\% and 4\%, respectively. The enrollment rate for middle-income families declined from 12\% in 1968 to 6\% in 2013. This decline was driven by shrinking Catholic school enrollment, which represented 89\% of private school enrollment in 1965 and 42\% of enrollment in 2013. Moreover, contemporary Catholic school students tend to be more affluent compared to prior decades. There has been a long-term increase in private school enrollment in the South, with middle-income families increasingly attending Christian schools and high-income families increasingly attending nonsectarian schools. With respect to race, 11\% of white students attended private school in 2013, compared to 5\% of Black students and 3\% of Hispanic students.

@@

The High School Longitudinal Survey of 2009 (HSLS) tracks the postsecondary destinations of students who were in ninth grade in 2009. Simple descriptive statistics are eye-opening (Author calculations).\footnote{Analyses consist of students who met the following conditions: completed the 2012 follow-up survey; obtained high school transcript data; and completed the 2016 follow-up survey. The survey weight variable used was W4W1W2W3STU. After weighting, these 14,020 (unweighted) students represent a population of 3.93 million U.S. ninth graders in 2009. Data on college rankings come from the 2020 U.S. News \& World Report \emph{Best Colleges}. Data on high school control come from the HSLS school file and from the Private School Survey.} About 93\% of students attended a public high school and 7\% attended a private high school, including 3.6\% attending a Catholic school, 1.6\% attending a nonsectarian school, and 1.8\% attending a Christian school. Of public high school students, 5.0\% attending a public university ranked in the top 100 by U.S. News and World Report and 1.9\% attended a private university ranked in the top 100. Of private high school students, 10.9\% attended a top 100 public university and 16.8\% attended a top 100 private university. For Catholic school students, 12.8\% attended a top 100 public university and 11.7\% attended a top 100 private university, while for Christian schools these percentages were 8.2\% and 8.2\%, respectively. However, for nonsectarian private school students, 11.3\% attended a top 100 public university and an incredible 39.8\% attended a top 100 private university.

Focusing on the institutional composition of private universities in U.S. News and World Report's top 100, 39.8\% attended a private high school (compared to 7\% of all high school students who attended a private high school) and 21.2\% attended a nonsectarian private high school (compared to 1.6\% of all high school students who attended a nonsectarian private high school).

@@

Analyses utilized several secondary data sources. The data on university characteristics come from the 2017 Integrated Postsecondary Education Data System (IPEDS) and 2020 Best Colleges ranking by U.S. News and World Report. For the high schools in our analysis, we use secondary data from the 2017-18 NCES Common Core of Data (CCD) for public high schools, and the 2017-18 NCES Private School Universe Survey (PSS) and 2020 Best Private High Schools ranking by Niche for private high schools.\footnote{For both public and private high schools, we also used NCES data from previous years when the 2017-18 data was not available for a small number of schools.}

@@

This article faces several limitations. First, our data collection sample consisted of public research-extensive universities as defined by the 2000 Carnegie Classification (N=102) and all private universities in the top 100 of U.S. News and World Report National Universities rankings (N=57). Making comparisons between public and private universities would be more appropriate if we used the same classification method to define the data collection sample for all universities. The process we used to define the data collection sample yielded a sample of public research universities with lower U.S. News and World Report rankings than the analysis sample of selective private universities. However, the columns in Table \ref{tab:sample-characteristics} for 25th and 75th percentile SAT score of enrolled freshmen suggest that private universities ranked lower than 75 have test score ranges that are fairly similar to public research universities not ranked in the top 100.
